### 2018-01-13 paper/jfp-submission/safe.tex
minor formatting

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, and @x@ and @y@
% WRW thinks "morally" is too informal for this venue. 
were \emph{originally} holes, we can treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

@@

Thus, it is not particularly satisfying that \toolname fails to produce
a witness here; a possible solution could be to track the types of the
inputs, and demonstrate to the user how they change between recursive
calls.
%
This would require maintaining a typing environment of variables in
addition to the environments we maintain for holes.
%
We would have to modify the rule $\reappgood$ from
Figure~\ref{fig:operational} to additionally $\forcesym$ the function's
type against the concrete inputs.
%
However, we would want to ensure that this $\forcesym$ cannot fail ---
it is preferable to report a stuck term as that provides a fuller view
of the error.
%
Rather, we would note which evaluation steps induced incompatible
type refinements, and if a traditional witness cannot be found, we could
then report a trace expanded to show precisely these steps.
%
This represents only a modest extension to our semantics, and would be
interesting to explore further.

The $\tfun$ abstraction is also problematic when we have to generate new
functions. Consider the following @pipe@ function that composes a list
of functions.
%
\begin{code}
  let pipe fs =
    let f a x = a x in
    let base a = a in
    List.fold_left f base fs
\end{code}
%
In the folding function @f@, the student has applied the accumulator @a@
to the new function @x@, rather than composing the two.
%
The \ocaml compiler again detects a cyclic typing constraint and rejects
the program, but \toolname is unable to produce a witness.
%
In this case the issue lies in the fact that the safety of a call to
@pipe@ is determined by its arguments: the call @pipe [(fun x -> 1)]@ is
safe, but the call @pipe [(fun x -> 1); (fun x -> 2); (fun x -> 3)]@
will get stuck when we try to reduce @1 (fun x -> 3)@.
%
Unfortunately, \toolname is unable to synthesize such a witness because
of the abstraction to $\tfun$.
%
Specifically, this abstraction forces our hand inside $\gensym$; we do
not know the what the input and output types of the function should be,
so the only safe thing to do is to generate a function that accepts any
input and returns a value of a yet-to-be-determined type.
%
Thus, \toolname's lenient instantiation of holes prevents it from
discovering a witness for @pipe@.

>>>>

\newpage
Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, and @x@ and @y@
% WRW thinks "morally" is too informal for this venue. 
were \emph{originally} holes, we can treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

@@

Thus, it is not particularly satisfying that \toolname fails to produce
a witness here; one solution could be to track the types of the
inputs, and demonstrate to the user how they change between recursive
calls.
%
This would require maintaining a typing environment of variables in
addition to the environments we maintain for holes.
%
We would have to modify the rule $\reappgood$ from
Figure~\ref{fig:operational} to additionally $\forcesym$ the function's
type against the concrete inputs.
%
However, we would want to ensure that this $\forcesym$ cannot fail ---
it is preferable to report a stuck term as that provides a fuller view
of the error.
%
Rather, we would note which evaluation steps induced incompatible
type refinements, and if a traditional witness cannot be found, we could
then report a trace expanded to show precisely these steps.
%
This represents only a modest extension to our semantics and would be
interesting to explore further.

The $\tfun$ abstraction is also problematic when we have to generate new
functions. Consider the following @pipe@ function that composes a list
of functions.
%
\begin{code}
  let pipe fs =
    let f a x = a x in
    let base a = a in
    List.fold_left f base fs
\end{code}
%
In the folding function @f@, the student has applied the accumulator @a@
to the new function @x@, rather than composing the two.
%
The \ocaml compiler again detects a cyclic typing constraint and rejects
the program, but \toolname is unable to produce a witness.
%
In this case the issue lies in the fact that the safety of a call to
@pipe@ is determined by its arguments: the call @pipe [(fun x -> 1)]@ is
safe, but the call @pipe [(fun x -> 1); (fun x -> 2); (fun x -> 3)]@
will get stuck when we try to reduce @1 (fun x -> 3)@.
%
Unfortunately, \toolname is unable to synthesize such a witness because
of the abstraction to $\tfun$.
%
Specifically, this abstraction forces our hand inside $\gensym$; we do
not know the what the input and output types of the function should be,
so the only safe thing to do is to generate a function that accepts any
input and returns a value of a yet-to-be-determined type.
%
Thus, our lenient instantiation of holes prevents us from
discovering a witness here.
